---
title: "Untitled"
format: html
server: shiny
---

## Shiny Documents

This Quarto document is made interactive using Shiny. Interactive documents allow readers to modify parameters and see the results immediately. Learn more about Shiny interactive documents at <https://quarto.org/docs/interactive/shiny/>.

## Inputs and OutputsS

You can embed Shiny inputs and outputs in your document. Outputs are automatically updated whenever inputs change. This demonstrates how a standard R plot can be made interactive:'

```{r}
#| context: setup
#| include: false

library(sf)
library(rdeck)

hex_grid <- read_sf('sf_input/melbourne_hex_grid.gpkg')
mesh_block_sf <- read_sf('sf_input/mb_sf.gpkg')


```

```{r}
#| panel: sidebar
actionButton("toggle_layer", "Switch Layer", class = "btn-primary")
```

```{r}
rdeckOutput("map", width = "100%", height = "500px")
```

```{r}
#| context: server
output$map <- renderRdeck({
  
   # React to button press
  layer_type <- input$toggle_layer %% 2
  
  base_map <- rdeck(initial_bounds = st_bbox(mesh_block_sf))
    
  
  if (layer_type == 0) {
    # Layer 1: Points
    base_map %>%
      add_polygon_layer(
        data = mesh_block_sf,
        get_polygon = geom,
        get_fill_color = scale_color_quantile(value_column, palette = scales::viridis_pal(), probs = seq.int(0,1,0.1)),
        pickable = TRUE
      )
  } else {
    # Layer 2: Polygons
    base_map %>%
      add_polygon_layer(opacity = 0.5,
        data = hex_grid,
        get_polygon = geom,
        get_fill_color = scale_color_quantile(max_employment, palette = scales::viridis_pal(), probs = seq.int(0,1,0.1)),
        pickable = TRUE
      )
  }
  

})
```
